---
title: "When to use incomeLato, and when not to"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When to use incomeLato, and when not to}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The Latinobarómetro asks about political attitudes and behavior but has no
income measure that is comparable across countries and waves. `incomeLato`
fills that gap with a probability: for each respondent, how likely they are
to be in the bottom 50%, the middle 40% or the top 10% of their own country
in that survey year. Those probabilities are good at one job and bad at
another, and this vignette is about telling the two apart.

## Good use: aggregate composition

Use the probabilities as continuous weights to describe groups. The question
"what share of the poorest half of Brazilians supported party X in 2018?" is
answered by a weighted mean of the party indicator, with weight
`prob_bottom50 * survey_weight`. In LAPOP, which records household income,
the same shares can be computed twice, weighting respondents once by their
observed income group and once by the predicted probabilities. With the
packaged model, the two agree closely (357 country-wave-party comparisons,
15 countries, 2008–2018):

- a mean absolute difference of 0.71 percentage points in the bottom 50's
  support share;
- a Spearman correlation of 0.97–0.99 between the two sets of shares, across
  the three income groups;
- the party most supported by the bottom 50% is the same in 97% of 66
  country-waves.

## Bad use: classifying individuals

At the level of one respondent the model is weak. The Spearman correlation
between a person's predicted probability and their observed income group,
taken from the household income bracket they report to LAPOP, is 0.44 for
the bottom 50, 0.25 for the middle 40 and 0.35 for the top 10. Five
predictors (sex, age, education, car and washing machine ownership) cannot
place an individual reliably. So:

- do not turn the probabilities into a label (`prob_bottom50 > 0.5` → "poor")
  and treat that label as observed income;
- do not use the probabilities as an individual-level covariate whose
  coefficient you interpret as the effect of income.

## Comparing waves: yes. The size of the bottom 50% over time: no

Comparing groups across waves is a good use. Support for party X among the
bottom 50% in each wave, 1995–2023, is the same weighted mean computed wave
by wave: the probabilities describe the composition of the group in every
wave, and those compositions can be compared.

What cannot be read over time is the weight itself. `mean(prob_bottom50)`
per country-wave is not an estimate of how large the bottom half is, and
its change over time is not a trend: the aggregate margin drifts for
reasons internal to the model. The column `veredito_deriva_temporal` gives a
per-country drift label; its magnitude categories come from the
leave-one-wave-out validation of an earlier model vintage (v1.1).
`flag_margin` marks the countries where the margin did not validate cleanly
(`NA` for Mexico, where it was never assessed). It remains a justified
caution, because the direction of the drift persists under the packaged
model.

## Validation in detail

The package applies the v1.5 model with the v1.4 recalibrators (Mexico
uncalibrated). The table compares that packaged recipe with the two models
without recalibration, against LAPOP observed income. For the recipe, the
check is in-sample for the recalibrators, which were fitted on this same
LAPOP data; LAPOP never enters the training of the model itself.

| | Packaged recipe | v1.5, without recalibration | v1.4, without recalibration |
|---|---|---|---|
| Country-waves (countries) | 66 (15) | 66 (15) | 62 (14) |
| Share of the bottom 50%, predicted minus observed: median absolute difference | 2.3 pp | 7.3 pp | 8.0 pp |
| Same, largest absolute difference | 11.6 pp | 25.9 pp | 26.8 pp |
| Individual Spearman, bottom 50 / middle 40 / top 10 | 0.44 / 0.25 / 0.35 | 0.42 / 0.23 / 0.35 | 0.42 / 0.21 / 0.35 |
| Party support, mean absolute difference, bottom 50 | 0.71 pp | 0.67 pp | 0.68 pp |
| Party support, Spearman across the three groups | 0.97–0.99 | 0.97–0.99 | 0.97–0.99 |
| Most-supported party of the bottom 50% matches | 97% | 97% | 98% |

Recalibration fixes the level of each income group. It barely changes the
composition of party support, because support within a group is a ratio of
weighted sums, and a shift in the level of the weights largely cancels out.

Under the recipe, 11 of 15 countries have a median difference within ±1 pp;
Ecuador (−2.1), Uruguay (−2.5) and Mexico (−2.7, no recalibrator) fall
outside ±2 pp. Single country-waves vary more: the largest difference is
above 2 pp in 14 of the 15 countries.

## A minimal example

The example follows the normal path: a wave written to a file, then
`income_lato()`. The wave is synthetic, generated by the package; no real
respondent is involved. `supports_democracy` is **simulated** (1 = democracy
is preferable to any other kind of government), with a probability that
rises with predicted income, so that the table shows a gradient.

```{r}
library(incomeLato)
sim <- simulate_latinobarometro_wave(2011, n = 200)
arquivo <- file.path(tempdir(), "Latinobarometro_2011_sim.dta")
haven::write_dta(sim, arquivo)

res <- income_lato(arquivo)
res <- res[res$elegivel, ]   # scored respondents only

# simulated outcome, rising with predicted income
set.seed(1)
p <- 0.40 + 0.25 * res$prob_middle40 + 1.2 * res$prob_top10
res$supports_democracy <- rbinom(nrow(res), 1, pmin(p, 1))

# share supporting democracy in each income group, by country:
# each respondent weighted by prob_<group> times the survey weight
grupos <- c("bottom50", "middle40", "top10")
share <- sapply(grupos, function(g) {
  w <- res[[paste0("prob_", g)]] * res$weight
  tapply(w * res$supports_democracy, res$iso3, sum) / tapply(w, res$iso3, sum)
})
round(100 * share, 1)
```

```{r, include = FALSE}
unlink(arquivo)
```

With real data, the variable comes from the Latinobarómetro's standard
question on preference for democracy. Its name changes between waves, so
look it up in each wave's questionnaire. `income_lato()` does not carry over
the other columns of your file; bring the variable in with
`raw$variable[res$respondent_row_id]`, where `raw` is the wave read with
`haven::read_dta()` and `respondent_row_id` is the row's position in the
file.
