---
title: "When to use incomeLato, and when not to"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When to use incomeLato, and when not to}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The Latinobarómetro asks about political attitudes and behavior but has no
income measure that is comparable across countries and waves. `incomeLato`
fills that gap with a probability: for each respondent, how likely they are
to be in the bottom 50%, the middle 40% or the top 10% of their own country
in that survey year. Those probabilities are good at one job and bad at
another, and this vignette is about telling the two apart.

## Good use: aggregate composition

Use `prob_bottom50` as a continuous weight to describe groups. The question
"what share of the poorest half of Brazilians supported party X in 2018?" is
answered by a weighted mean of the party indicator, with weight
`prob_bottom50 * survey_weight`. In LAPOP, which records household income,
the same shares can be computed twice, weighting respondents once by their
observed income group and once by the predicted probabilities. The two
agree closely:

- Spearman correlation of 0.97–0.99 between the two sets of shares, across
  the three income groups (339 country-wave-party comparisons, model v1.4);
- under the current model v1.5, a mean absolute difference of 0.670
  percentage points in the bottom-50 support share (357 country-wave-party
  comparisons, 15 countries), with the party most supported by the bottom
  50% the same in 97% of 66 country-waves.

## Bad use: classifying individuals

At the level of one respondent the model is weak: Spearman correlations of
0.21–0.42 between a person's predicted probability and their observed
income group, taken from the household income bracket they report to LAPOP
(92,940 respondents, 14 countries, 2008–2023, validation under v1.4). Five predictors
(sex, age, education, car and washing machine ownership) cannot place an
individual reliably. So:

- do not turn the probabilities into a label (`prob_bottom50 > 0.5` → "poor")
  and treat that label as observed income;
- do not use the probabilities as an individual-level covariate whose
  coefficient you interpret as the effect of income.

## Comparing waves: yes. The size of the bottom 50% over time: no

Comparing groups across waves is a good use. Support for party X among the
bottom 50% in each wave, 1995–2023, is the same weighted mean computed wave
by wave: the probabilities describe the composition of the group in every
wave, and those compositions can be compared.

What cannot be read over time is the weight itself. `mean(prob_bottom50)`
per country-wave is not an estimate of how large the bottom half is, and
its change over time is not a trend: the aggregate margin drifts for
reasons internal to the model. The column
`veredito_deriva_temporal` reports the drift verdict per country from the
article's validation, and `flag_margin` marks the countries where the margin
did not validate cleanly (`NA` for Mexico, where it was never assessed).

## A minimal example

The package includes a generator of synthetic data with the structure of a
Latinobarómetro wave; no real respondent is involved.

```{r}
library(incomeLato)
sim <- simulate_latinobarometro_wave(2018, n = 50)
core <- harmonize_core_predictors(sim, 2018)
educ <- harmonize_educ3_latinobarometro(sim, 2018)
cw <- read.csv(system.file("extdata", "country_crosswalk.csv", package = "incomeLato"))
banco <- data.frame(year = 2018, country_raw = core$country_raw,
                    respondent_row_id = seq_len(nrow(sim)), weight = core$weight,
                    sexo_final = core$sexo_final, idade_final = core$idade_final,
                    educ3_final = educ$educ3_final, owns_car = core$owns_car,
                    owns_washing_machine = core$owns_washing_machine)
banco <- merge(banco, cw, by.x = "country_raw", by.y = "country_numeric")
res <- suppressWarnings(predict_latinobarometro_income(banco))

# aggregate use: support for a (synthetic) party among the bottom 50%,
# weighting each respondent by prob_bottom50 times the survey weight
set.seed(1)
res$supports_A <- rbinom(nrow(res), 1, 0.4)
ok <- !is.na(res$prob_bottom50)
w <- res$prob_bottom50[ok] * res$weight[ok]
round(tapply(w * res$supports_A[ok], res$iso3[ok], sum) / tapply(w, res$iso3[ok], sum), 2)
```

With real data, replace the synthetic steps with a single call:
`income_lato("folder/with/latinobarometro/files")`.
