---
title: "When to use incomeLato, and when not to"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When to use incomeLato, and when not to}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The Latinobarómetro asks about political attitudes and behavior but has no
income measure that is comparable across countries and waves. `incomeLato`
fills that gap with a probability: for each respondent, how likely they are
to be in the bottom 50%, the middle 40% or the top 10% of their own country
in that survey year. Those probabilities are good at one job and bad at
another, and this vignette is about telling the two apart.

## Good use: aggregate composition

Use `prob_bottom50` as a continuous weight to describe groups. The question
"what share of the poorest half of Brazilians supported party X in 2018?" is
answered by a weighted mean of the party indicator, with weight
`prob_bottom50 * survey_weight`. Validated against LAPOP, this kind of
aggregate recovers the observed shares closely:

- Spearman correlation between predicted and observed shares of 0.97–0.99
  across the three income groups (339 country-wave-party comparisons,
  per-class validation under model v1.4);
- under the current model v1.5, a mean absolute error of 0.670 percentage
  points in the bottom-50 support share, with the most-supported party
  matching in 97% of 357 comparisons across 15 countries.

## Bad use: classifying individuals

At the level of one respondent the model is weak: Spearman correlations of
0.21–0.42 between a person's predicted probability and their observed
position (92,940 LAPOP respondents, validation under v1.4). Five predictors
(sex, age, education, car and washing machine ownership) cannot place an
individual reliably. So:

- do not turn the probabilities into a label (`prob_bottom50 > 0.5` → "poor")
  and treat that label as observed income;
- do not use the probabilities as an individual-level covariate whose
  coefficient you interpret as the effect of income.

## Also not: the size of the bottom 50% over time

`mean(prob_bottom50)` per country-wave is not an estimate of how large the
bottom half is, and its movement over time is not a trend. The aggregate
margin drifts for reasons internal to the model. The column
`veredito_deriva_temporal` reports the drift verdict per country from the
article's validation, and `flag_margin` marks the countries where the margin
did not validate cleanly (`NA` for Mexico, where it was never assessed).

## A minimal example

The package includes a generator of synthetic data with the structure of a
Latinobarómetro wave; no real respondent is involved.

```{r}
library(incomeLato)
sim <- simulate_latinobarometro_wave(2018, n = 50)
core <- harmonize_core_predictors(sim, 2018)
educ <- harmonize_educ3_latinobarometro(sim, 2018)
cw <- read.csv(system.file("extdata", "country_crosswalk.csv", package = "incomeLato"))
banco <- data.frame(year = 2018, country_raw = core$country_raw,
                    respondent_row_id = seq_len(nrow(sim)), weight = core$weight,
                    sexo_final = core$sexo_final, idade_final = core$idade_final,
                    educ3_final = educ$educ3_final, owns_car = core$owns_car,
                    owns_washing_machine = core$owns_washing_machine)
banco <- merge(banco, cw, by.x = "country_raw", by.y = "country_numeric")
res <- suppressWarnings(predict_latinobarometro_income(banco))

# aggregate use: support for a (synthetic) party among the bottom 50%,
# weighting each respondent by prob_bottom50 times the survey weight
set.seed(1)
res$supports_A <- rbinom(nrow(res), 1, 0.4)
ok <- !is.na(res$prob_bottom50)
w <- res$prob_bottom50[ok] * res$weight[ok]
round(tapply(w * res$supports_A[ok], res$iso3[ok], sum) / tapply(w, res$iso3[ok], sum), 2)
```

With real data, replace the synthetic steps with a single call:
`income_lato("folder/with/latinobarometro/files")`.
